---
title: "Statistical Analysis in Political Science II:\\newline More issues in modeling time series"
author: "JBrandon Duck-Mayr"
date: "`r quack::american_date_format(Sys.Date())`"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}{\setlength{\itemsep}{3pt}}
    - \newcommand{\setskip}{\setlength{\parskip}{3pt}}
    - \renewcommand{\tightlist}{\setsep\setskip}
    - \newcommand{\expectation}[1]{\ensuremath{\mathbb{E}\left[#1\right]}}
    - \newcommand{\pd}[2][]{\ensuremath{\frac{\partial{#1}}{\partial{#2}}}}
    - \DeclareMathOperator{\Var}{Var}
    - \DeclareMathOperator{\Cov}{Cov}
    - \newcommand{\variance}[1]{\ensuremath{\Var\left[#1\right]}}
    - \newcommand{\covariance}[1]{\ensuremath{\Cov\left[#1\right]}}
    - \DeclareMathOperator{\standarddeviation}{sd}
    - \DeclareMathOperator{\standarderror}{se}
    - \usepackage{siunitx}
output:
    quack::presentation:
        toc: false
        incremental: true
---

<!--
Topics:
* Prais-Winsten
* Cox Proportional Hazards
-->

## Time series issues and topics

. . .

- Linear regression for time series
  + Revised assumptions, stationarity, types of exogeneity, ...
  + Lagged predictors or outcomes
  + Serial autocorrelation, HAC standard errors
  + Prais-Winsten estimation
- Time series cross-sectional data
- Survival analysis

## Prais-Winsten estimation

. . .

- The problem: HAC standard errors can be unreasonably large
- A solution(?): Prais-Winsten estimation
  + A form of "feasible generalized least squares" (feasible GLS)
  + **Warning**: Requires *strict* exogeneity for consistency!!
- Start by assuming TS1--TS4 & that errors follow an AR(1) model,
  $$ u_t = \rho u_{t-1} + \varepsilon_t $$
- Consider the equation $$ y_t = \beta_0 + \beta_1 x_{t1} + \dots + \beta_k x_{tk} + u_t; $$ multiply the previous observation by $\rho$ & subtract it to get $$ \tilde{y}_t = (1 - \rho) \beta_0 + \beta_1 \tilde{x}_{t1} + \dots + \beta_k \tilde{x}_{tk} + \varepsilon_t $$ where $$ \tilde{y}_t = y_t - \rho y_{t-1}, \tilde{x}_{tk} = x_{tk} - \rho x_{(t-1)k} $$ are the **quasi-differenced data**.

## Prais-Winsten estimation

. . .

- Note that this means we have to drop the first observation\invisible<1-2>{,}
- unless we define $$ \tilde{y}_1 = (1 - \rho^2)^{1/2} \beta_0 + \beta_1 \tilde{x}_{11} + \dots + \beta_k \tilde{x}_{1k} + \tilde{\varepsilon}_1, $$ $$ \text{where } \tilde{Z}_1 = (1 - \rho^2)^{1/2} Z_1 $$
- Then $t$ and $F$ statistics are asymptotically valid
- **but**, we don't know $\rho$...
- but luckily using $\hat{\rho}$ doesn't affect the estimators' asymptotic distribution

## Prais-Winsten estimation

. . .

So, the procedure is

1. Run OLS of $y_t$ on $x_{t1}, \dots x_{tk}$ with residuals $\hat{u}_t$
2. Regress $\hat{u}_t$ on $\hat{u}_{t-1}$ to get $\hat{\rho}$
3. Run OLS on the quasi-differenced data
   - Then your $t$ and $F$ statistics are asymptotically valid
   - (assuming TS1--TS4 and strict exogeneity hold)
   - Your standard errors will be more efficient,
     but at the cost of stronger assumptions

## Survival analysis

. . .

- Sometimes we care most about *when* something happens
  + How long does it take to confirm a Supreme Court justice?
  + Why do military conflicts persist, or not?
- Some important terms:
  + Survival function: Probability of an event *not* happening before or at time $t$
  + Hazard function: Instantaneous risk of the event happening
  + Cumulative hazard function: Integral of hazard function

## Proportional Hazards

. . .

- Consider two hazard functions, $h_0\left(t\right)$ and $h_1\left(t\right)$
- If $h_1\left(t\right) = \psi h_0\left(t\right)$, for some positive constant $\psi$ and all $t \geq 0$, we say they are **proportional** hazard functions

## Log-rank test

. . .

We can test equality of survival functions with a **log-rank test**

```{r log-rank-example}
## (see Table 3.3 in Broström 2012)
exdata = data.frame(
    group = rep(x = c("numbers", "letters"), each = 5),
    time  = c(4, 2, 6, 1, 3.5, 5, 3, 6, 1, 2.5),
    event = c(1, 0, 1, 1, 0, 1, 1, 0, 1, 0)
)
library(survival)
fit = coxph(Surv(time, event) ~ group, data = exdata)
summary(fit)$logtest
```

## Visually assessing proportional hazards

. . .

```{r proportional-example-plot, warning=FALSE, height = 2, width = 3, out.height = "50%", fig.align = "center"}
library(ggplot2)
library(ggfortify)
fit = survfit(
    Surv(enter, exit, event) ~ sex,
    data = eha::oldmort
)
autoplot(fit, xlim = c(60, NA))
```

## Building the Cox proportional hazards model

. . .

- Let $\beta = \log\left(\psi\right)$. Then $$ h_x\left(t\right) = \exp\left(\beta x\right) h_0\left(t\right) $$
- Now consider a data structure where each observation $i$ has $(t_{i0}, t_{i}, d_{i}, \mathbf{x}_i)$, representing the starting point, ending point, event indicator, and predictors respectively
- Model: $(t_{i0}, t_{i}, d_{i})$ is the response
- Interpretation: $$ \frac{h\left(t; x+1\right)}{h\left(t; x\right)} = \exp\left(\beta\right) $$
- Implementation:
  + Create a survival object; e.g. `Surv(time, event)`
  + Use `survival::coxph()`; e.g. `coxph(Surv(time, event) ~ group, data = exdata)`
  + Then there are various summary and plotting methods, demonstrated above

## Cox Proportional Hazards `summary()` example

\footnotesize

```{r coxph-summary-example}
fit = coxph(Surv(enter, exit, event) ~ sex, data = eha::oldmort)
summary(fit)
```

## Time series cross-sectional data

. . .

- A full treatment of this topic is outside the scope of this course
- But this is an important type of data analysis in political science
- Often used for causal inference difference-in-differences design
  + **Warning**: There have been a lot of developments on this "recently"
  + One resource on that for the interested is Ruttenauer and Aksoy (2024): <https://arxiv.org/pdf/2402.09928.pdf>

## Some considerations

. . .

- unit effects
- panel-corrected standard errors
  + `psce::psce(lm(...))`
- lagged DV
- again, be wary of recent developments !!!
